Metropolitan Areas (Deaths & Case)

### 2020-11-12 KPMG Team/Dashboard/visualization_scripts/msa_county_state_cities.Rmd

@@

# plotting 
library(ggplot2)
library(hrbrthemes) 
library(geofacet)

@@

# metropolitan state areas delineation from file taken off Census Bureau
data <- read_csv("msa.csv")

#nytimes county-level cases & deaths data from online link 
nyt <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

# census data API, variables can be changed here 
census <- get_acs(geography = "county",
                   variables = c(population = "B01003_001E"), # total population), # black veterans 
                   output = "wide",
                   county = "",
                   year = 2018)

city <- read_csv("cities.csv")

View(city)
```

@@

# --- *** --- metropolitan statistical areas  --- *** --- 
# comprised of counties 
# is a U.S. geographic area defined by the Office of Management and Budget (OMB) that consists of one or more counties (or equivalents) anchored by an urban center of at least 10,000 people plus adjacent counties that are socioeconomically tied to the urban center by commuting.
# https://en.wikipedia.org/wiki/List_of_metropolitan_statistical_areas

@@

# list of top msas by population, if we want
# top_msa_popn <- census_msa_merge %>%
#   select(msa, msa_title, msa_population) %>%
#   arrange(desc(msa_population)) %>%
#   distinct() %>%
#   na.omit() %>%
#   data.frame() %>%
#   slice(1:12)

# census_msa <- census_msa_merge %>%
#   semi_join(top_msa_popn, by="msa")

@@

# ------------------------------------ #
names(nyt)

county_sum <- nyt %>%
  filter(date == max(nyt$date)) # latest

# only need county to merge both datasets, will keep just metro areas (csa)
msa_covid_merge <- census_msa %>% 
  left_join(nyt, by = "fips") %>% 
  select(msa, msa_title, msa_population, date, cases, deaths) %>% 
  filter(date >= lubridate::as_date("2020-01-24") & 
           date <= lubridate::as_date(Sys.Date() - 8)) # 8 days prior, so that the 7 day mean wouldn't get messed up

# get 7-day mean of cases with zoo::rolling mean and dplyr::lag function 
msa_covid_cases <- data.frame(msa_covid_merge) %>% 
  group_by(msa, date) %>% 
  dplyr::summarise(cases_sum = sum(cases, na.rm = TRUE)) %>% 
  mutate(new_cases = cases_sum - dplyr::lag(x = cases_sum, n = 1, order_by = date)) %>%
  mutate(new_cases_mean = rollmean(new_cases, k = 7, fill = NA))

# same as above, but for deaths
msa_covid_deaths <- data.frame(msa_covid_merge) %>% 
  group_by(msa, date) %>% 
  dplyr::summarise(deaths_sum = sum(deaths, na.rm = TRUE)) %>% 
  mutate(new_deaths = deaths_sum - dplyr::lag(x = deaths_sum, n = 1, order_by = date)) %>%
  mutate(new_deaths_mean = rollmean(new_deaths, k = 7, fill = NA))

# join deaths and cases numbers together (not sure if I can do it in one?)
# contains
msa_covid <- msa_covid_merge %>% 
  left_join(msa_covid_cases, by = c("msa", "date")) %>% 
  left_join(msa_covid_deaths, by = c("msa", "date")) %>% 
  arrange(desc(msa_population))

# final count of msa cases, in order to get ranking
msa_covid_sum <- msa_covid %>% 
  group_by(msa, msa_title, msa_population) %>%
  mutate(total_cases = sum(new_cases)) %>% 
  mutate(total_deaths = sum(new_deaths)) #%>%
  arrange(desc(total_cases)) %>% 
  select(msa, msa_title, msa_population, total_cases, total_deaths) %>% 
  distinct() %>% # remove repeated values 
  na.omit()
  # drop_na(msa) # remove rows with NA in msa column

msa_covid_sum <- msa_covid %>% 
  filter(date == max(msa_covid$date)) %>% 
  select(msa, msa_title, msa_population, date, cases_sum, deaths_sum) #%>% 
  # distinct() #%>%
  # drop_na(msa)

View(msa_covid_sum)

table(msa_covid$msa_title)

@@

# similarly, calculate rolling daily 7-day average separately for cases
nyt_state_cases <- nyt %>% 
  group_by(state, date) %>%
  dplyr::summarise(cases_sum = sum(cases, na.rm = TRUE)) %>% 
  mutate(new_cases_state = cases_sum - dplyr::lag(x = cases_sum, n = 1, order_by = date)) %>%
  mutate(new_cases_state_mean = rollmean(new_cases_state, k = 7, fill = NA))

# rolling daily 7-day average for deaths
nyt_state_deaths <- nyt %>% 
  group_by(state, date) %>%
  dplyr::summarise(deaths_sum = sum(deaths, na.rm = TRUE)) %>% 
  mutate(new_deaths_state = deaths_sum - dplyr::lag(x = deaths_sum, n = 1, order_by = date)) %>%
  mutate(new_deaths_state_mean = rollmean(new_deaths_state, k = 7, fill = NA))

# combined total figures
state_covid <- nyt_state_cases %>% 
  left_join(nyt_state_deaths, by = c("state", "date")) %>%
  filter(date >= lubridate::as_date("2020-01-24") & 
         date <= lubridate::as_date(Sys.Date() - 8))

@@

# -------------------- FUNCTION TO MAKE A PLOT FOR ONE STATE  ------------------- #

stateData <- function(state_name) {
  state_covid %>% 
    filter(state == state_name)
}

metroData <- function(msa_id) {
  msa_covid %>% 
    filter(msa == msa_id)
}

covidCasesPlot <- function(state_data, state_name, metro_data) {

@@

`
```{r}

@@

# --------------------- TO MAKE A PLOT FOR ONE STATE -------------------------- #

@@

@@

# mapping packages needed 
library(leaflet)
library(maptools)
library(rgdal)

# interactive visualization
library(plotly)

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

library(tigris)
library(usmap) # to get US Map Data

@@

@@

outside_mainland <- c("02", "15", "72", "2", "3")
islands <- c("Virgin Islands", "Northern Mariana Islands")

@@

county_map_NY <- county_map %>% 
  filter(STATEFP == 36) # remove alaska and hawaii

# # plot for states 
# ggplot(data = county_map_NY) +
#   geom_sf(aes(fill = cases)) + 
#   scale_fill_viridis_c(option = "magma", direction = -1) +
# 
# 
# names(county_map_US)
# 
# fips(state = "New York")

# PLOT FUNCTION 
# county state map 
county_state_map <- function(state_name) {

@@

  ggplot(data = county_map_state) +
  geom_sf(aes(fill = cases)) + 
  scale_fill_viridis_c(option = "magma", direction = -1)

@@

W <- county_state_map("Wisconsin")

range(county_map_US$cases)

>>>>

# Data Manipulation

@@

# plotting 
library(ggplot2)
# library(hrbrthemes) 
# library(geofacet)

@@

# metropolitan state areas delineation from file taken off Census Bureau
data <- read_csv("data/msa.csv")

#nytimes county-level cases & deaths data from online link 
covid <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

# census data API, variables can be changed and added here 
census <- get_acs(geography = "county",
                   variables = c(population = "B01003_001E"), # total population
                   output = "wide",
                   county = "",
                   year = 2018) # ACS 2018 Survey

# US City & FIPS code boundaries 
# https://simplemaps.com/data/us-cities license 
city <- read_csv("data/uscities.csv")

# ---------------------------------------------------------- #
# -----------  DATA MANIPULATION FOR COUNTY  --------------- # 
# ---------------------------------------------------------- #

# not much to do
county_sum <- covid %>%
  filter(date == max(covid$date)) # latest

# Preview 
head(county_sum, 50)

@@

# --- *** --- metropolitan statistical areas  --- *** --- 
# only 31 MSA in this list
# comprised of counties 
# is a U.S. geographic area defined by the Office of Management and Budget (OMB) that consists of one or more counties (or equivalents) anchored by an urban center of at least 10,000 people plus adjacent counties that are socioeconomically tied to the urban center by commuting.
# https://en.wikipedia.org/wiki/List_of_metropolitan_statistical_areas

@@

# list of top msas by population, if we want
top_msa_popn <- census_msa_merge %>%
  select(msa, msa_title, msa_population) %>%
  arrange(desc(msa_population)) %>%
  distinct() %>%
  na.omit() %>%
  data.frame() %>%
  slice(1:12)

top_census_msa <- census_msa_merge %>%
  semi_join(top_msa_popn, by="msa")

@@

# ------------------------------------ #

# only need county to merge both datasets, will keep just metro areas (csa)
msa_covid_merge <- census_msa %>% 
  left_join(covid, by = "fips") %>%  # join nytimes county-covid df with population and msa ares
  select(msa, msa_title, msa_population, date, cases, deaths) %>% 
  # set date range
  filter(date >= lubridate::as_date("2020-01-24") &
           date <= lubridate::as_date(Sys.Date() - 8)) # 8 days prior, so that the 7 day mean wouldn't get messed up

# get 7-day mean of cases with zoo::rolling mean and dplyr::lag function 
msa_covid_cases <- data.frame(msa_covid_merge) %>% 
  group_by(msa, date) %>% 
  dplyr::summarise(cases_sum = sum(cases, na.rm = TRUE)) %>% 
  mutate(new_cases = cases_sum - dplyr::lag(x = cases_sum, n = 1, order_by = date)) %>% 
  # rollmean function from zoo package to help calculate rolling mean, k = no. of days
  mutate(new_cases_mean = zoo::rollmean(new_cases, k = 7, fill = NA))

# same as above, but for deaths
msa_covid_deaths <- data.frame(msa_covid_merge) %>% 
  group_by(msa, date) %>% 
  dplyr::summarise(deaths_sum = sum(deaths, na.rm = TRUE)) %>% 
  mutate(new_deaths = deaths_sum - dplyr::lag(x = deaths_sum, n = 1, order_by = date)) %>%
  # rollmean function from zoo package to help calculate rolling mean, k = no. of days
  mutate(new_deaths_mean = zoo::rollmean(new_deaths, k = 7, fill = NA))

# join deaths and cases numbers together (not sure if I can do it in one?)
# contains
msa_covid <- msa_covid_merge %>% 
  left_join(msa_covid_cases, by = c("msa", "date")) %>% 
  left_join(msa_covid_deaths, by = c("msa", "date")) %>% 
  arrange(desc(msa_population)) # arrange by msa with highest population

# final count of msa cases, in order to get ranking
msa_covid_sum <- msa_covid %>% 
  filter(date == max(msa_covid$date)) %>% 
  select(msa, msa_title, msa_population, date, cases_sum, deaths_sum) %>% 
  distinct() %>%
  drop_na(msa)

@@

# similarly, calculate rolling daily 7-day average separately for cases
covid_state_cases <- covid %>% # nyt dataset 
  group_by(state, date) %>%
  dplyr::summarise(cases_sum = sum(cases, na.rm = TRUE)) %>% 
  mutate(new_cases_state = cases_sum - dplyr::lag(x = cases_sum, n = 1, order_by = date)) %>%
  mutate(new_cases_state_mean = rollmean(new_cases_state, k = 7, fill = NA))

# rolling daily 7-day average for deaths
covid_state_deaths <- covid %>% 
  group_by(state, date) %>%
  dplyr::summarise(deaths_sum = sum(deaths, na.rm = TRUE)) %>% 
  mutate(new_deaths_state = deaths_sum - dplyr::lag(x = deaths_sum, n = 1, order_by = date)) %>%
  mutate(new_deaths_state_mean = rollmean(new_deaths_state, k = 7, fill = NA))

# combined total figures
state_covid <- covid_state_cases %>% 
  left_join(covid_state_deaths, by = c("state", "date")) %>%
  filter(date >= lubridate::as_date("2020-01-24") & 
         date <= lubridate::as_date(Sys.Date() - 8))

@@

# ------------------------------------------- #
# ----------- GETTING CITIES ---------------- # 
# ------------------------------------------- #

# cities are actually smaller than counties, e.g. New York City is part of New York County
# some are incorporated, and some would have mayors 
# this is useful data also to get labels and plot points

head(city)
```

# Visualizations

## Covid Line Graph (7-day average) per State

# \/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\\/\/\/ #
# *************** FUNCTION: PLOT DAILY COVID ONE STATE GRAPH ********************** # 
# \/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\\/\/\/ #

# plot line graph of daily cases for state and metropolitan area

# choose state name
stateData <- function(state_name) {
  state_covid %>% 
    filter(state == state_name)
}

# choose metropolitan state area ID 
# find them here: head(msa_covid_sum, 10)
metroData <- function(msa_id) {
  msa_covid %>% 
    filter(msa == msa_id)
}

# line graph
covidCasesPlot <- function(state_data, state_name, metro_data) {

@@

### Examples

@@

# ----------------- EG TO MAKE A DAILY CASES/ DEATHS PLOT FOR ONE STATE ------------------- #

@@

## Geo Mapping Visualizations

### County Cases

@@

library(sf)

library(usmap) # to get US Map Data

@@

library(cowplot)
library(gridExtra) # for grid plot

@@

# remove non-mainland US parts 
outside_mainland <- c("02", "15", "72", "2", "3")
islands <- c("Virgin Islands", "Northern Mariana Islands")

@@

# \/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\\/\/\/ #
# *************** FUNCTION: PLOT DAILY COVID ONE STATE GRAPH ********************** # 
# \/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\\/\/\/ #

# county state map 
county_state_map <- function(state_name) {

@@

  ggplot(data = county_map_state) +
    geom_sf(aes(fill = cases)) + 
    scale_fill_viridis_c(option = "magma", direction = -1) + 
    labs(title = paste(state_name))

@@

# e.g. 
TX <- county_state_map("Texas")
CA <- county_state_map("California")
FL <- county_state_map("Florida")
NY <- county_state_map("New York")
IL <- county_state_map("Illinois")
GA <- county_state_map("Georgia")

# TO DO: 
# Facet it

plot_grid(TX, CA, FL, NY, IL, GA, ncol = 2)

ggsave("county-map-cases.pdf")

```
### Metropolitan Cases

# in progress

# limit to state 
# view by state, won't be able to plot all 
# check out why there aren't all 31 MSAs

msa_map <- county_map_US %>% 
  inner_join(top_census_msa, by = "fips")

View(msa_map)

names(top_census_msa)

dim(top_census_msa)

  ggplot(data = msa_map) +
    geom_sf(aes(fill = cases)) + 
    scale_fill_viridis_c(option = "magma", direction = -1)

### 2020-11-12 KPMG Team/Visualizations/Louisa/Metropolitan Areas/msa_county_state_cities.Rmd

# list of top msas by population, if we want
top_msa_popn <- census_msa_merge %>%
  select(msa, msa_title, msa_population) %>%
  arrange(desc(msa_population)) %>%
  distinct() %>%
  na.omit() %>%
  data.frame() %>%
  slice(1:12)

top_census_msa <- census_msa_merge %>%
  semi_join(top_msa_popn, by="msa")

@@

### County Cases

@@

library(sf)

library(usmap) # to get US Map Data

@@

library(cowplot)
library(gridExtra) # for grid plot

@@

  ggplot(data = county_map_state) +
    geom_sf(aes(fill = cases)) + 
    scale_fill_viridis_c(option = "magma", direction = -1) + 
    labs(title = paste(state_name))

@@

# e.g. 
TX <- county_state_map("Texas")
CA <- county_state_map("California")
FL <- county_state_map("Florida")
NY <- county_state_map("New York")
IL <- county_state_map("Illinois")
GA <- county_state_map("Georgia")

# TO DO: 
# Facet it

plot_grid(TX, CA, FL, NY, IL, GA, ncol = 2)

ggsave("county-map-cases.pdf")

```
### Metropolitan Cases

# in progress

# limit to state 
# view by state, won't be able to plot all 
# check out why there aren't all 31 MSAs

msa_map <- county_map_US %>% 
  inner_join(top_census_msa, by = "fips")

View(msa_map)

names(top_census_msa)

dim(top_census_msa)

  ggplot(data = msa_map) +
    geom_sf(aes(fill = cases)) + 
    scale_fill_viridis_c(option = "magma", direction = -1)

>>>>

# list of top msas by population, if we want
# top_msa_popn <- census_msa_merge %>%
#   select(msa, msa_title, msa_population) %>%
#   arrange(desc(msa_population)) %>%
#   distinct() %>%
#   na.omit() %>%
#   data.frame() %>%
#   slice(1:12)

# census_msa <- census_msa_merge %>%
#   semi_join(top_msa_popn, by="msa")

@@

@@

# # mapping packages needed 
# library(leaflet)
# library(maptools)
# library(rgdal)
# 
# # interactive visualization
# library(plotly)

library(sf)
# library(rnaturalearth)
# library(rnaturalearthdata)

# library(tigris)
library(usmap) # to get US Map Data

@@

@@

  ggplot(data = county_map_state) +
  geom_sf(aes(fill = cases)) + 
  scale_fill_viridis_c(option = "magma", direction = -1)

@@

# e.g. 
W <- county_state_map("Wisconsin")
